---
title: "Cluster Analysis"
author: "Owen Jones"
date: "2018-12-01"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.path = "img/owen-cluster-analysis/"
)
```

## Cluster Analysis

Cluster Analysis is a statistical tool for identifying structure in data and identifying groups based on similarity. It is used in various fields and has become known as a central tool in unsupervised machine learning. There are numerous algorithms and this worksheet focuses on a couple of use cases. 

The main function we will use is the `hclust` function from the `stats` package. This package ships with base R, so you should have it available. 

```{r iris data}
data(iris)
summary(iris)
```

## Clustering using the iris dataset

The idea here is that we are experimenting with the clustering algorithm using data where we **do** know something about the structure. In this case, we know what the species are. Will cluster analysis find this structure? Let's plot stuff. 

```{r pressure, echo=FALSE, message=FALSE}
library(ggplot2)
library(ggpubr)
A <- ggplot(iris,aes(Sepal.Length, Sepal.Width,colour = Species)) + 
geom_point() + theme(legend.position="none")
B <- ggplot(iris,aes(Petal.Length, Petal.Width,colour = Species)) + 
geom_point() + theme(legend.position="none")
ggarrange(A,B)
```

You can see that there are some clustering here. Let's quantify it by first calculating a distance matrix (`dist`) and then using `hclust` to perform a hierarchical cluster analysis. You can then plot the result with `plot`.

```{r}
names(iris)
x <- iris[,3:4]
x <- dist(x)
clustResult <- hclust(x,method="average")
#clustResult <- cluster::agnes(x, stand=FALSE,method = "ward")

plot(clustResult)
```

Let's cut the tree where we find 3 clusters. You can do this with the `cutree` function.
```{r}
iris$clusterID <- cutree(clustResult, 3)
```

Then we can see how well our cluster analysis has recovered the actual species.

```{r}
table(iris$clusterID,iris$Species)
```

Try repeating this with the rest of the data (sepals), and with other algorithms (e.g. `method = "complete"`)

## Comparing...

```{r}
clustResult2 <- hclust(x,method="ward.D2")
```


```{r}
dendextend::tanglegram(as.dendrogram(clustResult),as.dendrogram(clustResult2))
```
