---
title: "Random Effects"
author: "Patrick Barks"
date: '2017-09-28'
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.path = "img/patrick-random-effects/",
  collapse = FALSE,
  message = FALSE,
  warning = FALSE
)
```

### Preliminaries

```{r}
# install.packages(c('dplyr', 'tibble', 'ggplot2', 'gridExtra', 'nlme', 'lme4', 'sjPlot'))
library(dplyr)
library(tibble)
library(ggplot2)
library(gridExtra)
library(nlme)
library(lme4)
library(sjPlot)
```

### Generate data with hierarchical structure

```{r, message=FALSE}
library(dplyr)
library(tibble)

# testing a hypothesis that fish body condition is associated with parasite load
# samples from multiple lakes, incl. male vs. female, of varying age
n_lakes <- 9
n_traps_per_lake <- 10
n_fish_per_trap <- 10
N <- n_lakes * n_traps_per_lake * n_fish_per_trap

sigma_fish <- 1.2
sigma_alpha_lake <- 2.2
sigma_beta_lake <- 0.5
sigma_age <- 2.7
sigma_trap_within_lake <- 1.7

b_int <- 27.2        # intercept
b_sex <- -2.5        # fish sex
b_age <- 0.4         # fish age
b_parasite <- -1.0   # fish parasite load

dat <- expand.grid(lake_id = paste('Lake', 1:n_lakes),
                   trap_id = rep(1:n_traps_per_lake, n_fish_per_trap),
                   stringsAsFactors = F) %>%
  as_tibble() %>%
  mutate(trap_id = paste('T', trap_id, as.numeric(as.factor(lake_id)), sep = '-')) %>%
  arrange(lake_id, trap_id) %>%
  mutate(sex = sample(c(0:1), N, replace = T),
         age = rpois(N, 3),
         parasite = rlnorm(N, 1.5),
         parasite_log = log(parasite))

lake_df <- tibble(lake_id = sort(unique(dat$lake_id)),
                  alpha_lake = rnorm(n_lakes, mean = 0, sd = sigma_alpha_lake),
                  beta_lake = rnorm(n_lakes, mean = 0, sd = sigma_beta_lake))

traps_df <- tibble(trap_id = sort(unique(dat$trap_id)),
                   alpha_trap = rnorm(n_lakes * n_traps_per_lake, mean = 0, sd = sigma_trap_within_lake))

dat <- dat %>%
  left_join(lake_df, by = 'lake_id') %>%
  left_join(traps_df, by = 'trap_id')

X <- as.matrix(cbind(int = 1, dat[,c('sex', 'age', 'parasite_log')]))
dat$condition <- X %*% c(b_int, b_sex, b_age, b_parasite) + dat$alpha_lake + dat$beta_lake * dat$parasite_log + rnorm(N, 0, sigma_fish)
```

### Visualize

```{r}
### visualize data
head(dat)

ggplot(dat, aes(x = parasite_log, y = condition)) +
  geom_point()
# plot(condition ~ parasite_log, dat)

ggplot(dat, aes(x = as.factor(sex), y = condition)) +
  geom_boxplot()
# boxplot(condition ~ sex, dat)

ggplot(dat, aes(x = lake_id, y = condition)) +
  geom_boxplot()
# boxplot(condition ~ lake_id, dat)
```

### Simple linear regression, body condition vs. log parasite load

```{r}
# fit model
mod1 <- lm(condition ~ parasite_log, data = dat)

# summarize
summary(mod1)
anova(mod1)

# check residuals
dat$resid1 <- resid(mod1)

ggplot(dat, aes(x = lake_id, y = resid1)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, linetype = 2)

# examine model fit with ggplot
ggplot(dat, aes(x = parasite_log, y = condition)) +
  geom_point(size = 2) +
  geom_smooth(method = 'lm')
```

### Add fixed effect intercepts for lake_id

```{r}
# fit model
mod2 <- lm(condition ~ lake_id + parasite_log, data = dat)

# summarize
summary(mod2)
anova(mod2)

# check residuals
dat$resid2 <- resid(mod2)

# residuals vs. lake
ggplot(dat, aes(x = lake_id, y = resid2)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, linetype = 2)

ggplot(dat, aes(x = age, y = resid2)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = 2)

# residuals vs. parasite_log within lake
ggplot(dat, aes(x = parasite_log, y = resid2)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_grid(sex ~ lake_id)
```

### Add fixed effect slopes for lake_id (with respect to parasite load)

```{r}
# fit model
mod3 <- lm(condition ~ parasite_log*lake_id, data = dat)

# summarize
summary(mod3)
anova(mod3)

# check residuals
dat$resid3 <- resid(mod3)
ggplot(dat, aes(x = lake_id, y = resid3)) + geom_boxplot()        # resid vs. lake
ggplot(dat, aes(x = as.factor(sex), y = resid3)) + geom_boxplot() # resid vs. sex
ggplot(dat, aes(x = age, y = resid3)) + geom_point()              # resid vs. age

# examine model fit with ggplot
ggplot(dat, aes(x = parasite_log, y = condition, col = lake_id)) +
  geom_point(size = 2) +
  geom_smooth(method = 'lm') +
  scale_color_brewer(palette = 'Dark2') +
  theme(legend.position = c(0.07, 0.17))
```


### Add in terms for fish age and sex

```{r}
# fit model
mod4 <- lm(condition ~ age + sex + parasite_log*lake_id, data = dat)

# summarize
summary(mod4)$coefficients
anova(mod4)

# check residuals
dat$resid4 <- resid(mod4)
ggplot(dat, aes(lake_id, resid4)) + geom_boxplot()  # resid vs. lake
ggplot(dat, aes(sex, resid4)) + geom_boxplot()      # resid vs. sex
ggplot(dat, aes(age, resid4)) + geom_point()        # resid vs. age
```

### Model as above, but with random intercepts for lake_id (i.e. mixed effect model)

```{r}
# fit model
mod5a <- lme(condition ~ age + sex + parasite_log, random = ~ 1 | lake_id, data = dat)
mod5b <- lmer(condition ~ age + sex + parasite_log + (1 | lake_id), data = dat)

# summarize nlme
summary(mod5a)
anova(mod5a)

# summarize lme4 (see: help('pvalues', package = 'lme4'))
summary(mod5b)
anova(mod5b)
drop1(mod5b, test = 'Chisq')

# check random effects
ranef(mod5a)

# examine distribution describing of random intercepts
curve(dnorm(x, mean = 27.385716, sd = 3.998522), from = 0, to = 60)

# plot effects
plot_model(mod5b, type = 're', vars = 'parasite_log')
plot_model(mod5b, type = 'slope', terms = 'parasite_log', show.ci = T, point.alpha = 1)
```

### Random intercepts and slopes for lake_id

```{r}
# fit model
mod6a <- lme(condition ~ age + sex + parasite_log, random = ~ parasite_log | lake_id, data = dat)
mod6b <- lmer(condition ~ age + sex + parasite_log + (parasite_log | lake_id), data = dat)

# summarize nlme
summary(mod6a)
anova(mod6a)

# check random effects
ranef(mod6a)

# examine distributions describing random intercepts and slopes
curve(dnorm(x, mean = 26.512292, sd = 2.912040), from = 0, to = 60)   # random intercepts
curve(dnorm(x, mean = -1.264316, sd = 2.343843), from = -15, to = 15) # random slopes

# plot effects
plot_model(mod6b, type = 're')
plot_model(mod6b, type = 'slope', terms = 'parasite_log')

# confidence intervals
confint(mod6b, parm = '(Intercept)')
confint(mod6b, parm = 'parasite_log')
```

